---
title: "Getting started with scHINT"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with scHINT}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(scHINT)
```

scHINT estimates how much of the variance of a gene's expression is explained by
genetics (or by a perturbation), and how much of that depends on the **cell state**
or **cell type**, using Haseman-Elston regression on pairs of observations.

The example data are fully simulated: `schint_example` (genotypes, a cell-level table
and a sample-level table) and `schint_pert_example` (perturb-seq-like cells).

```{r data}
data(schint_example)
ex <- schint_example
str(ex, max.level = 1)
head(ex$cell[, 1:9], 3)
ex$truth
```

## Cell-level data: G x cell state

`geno` is the donor x SNP dosage matrix of the cis-SNPs of the gene. By default one GRM
is built from all SNPs. `context` is the cell-state variable the genetic effect may
depend on (numeric = continuous, factor/character = categorical); `covariates` only
enter as main effects.

```{r cell}
fit <- schint_cell(ex$cell, y = "GENE_A", id = "donor", geno = ex$geno,
                   context = "state",
                   covariates = c("pc1", "pc2", "age", "sex", "batch", "subtype"),
                   jackknife = TRUE)
fit
```

* `G` - genetic variance independent of the cell state.
* `GxC` - genetic variance that changes along the context (here `state`).
* `G_total` - `G + GxC`; `I` - donor effect that is not genetic; `context`, `covariate` -
  main effects.

The per-component table is `fit$coefficients`. Variance components are on the scale of
standardized expression, so they are fractions of the variance. A gene without
genotype-by-state effects:

```{r cell-b}
schint_cell(ex$cell, "GENE_B", "donor", geno = ex$geno, context = "state")$summary
```

### Several GRMs, or pre-computed GRMs

```{r grm}
# split cis-SNPs into three MAF bins -> three GRMs
fit3 <- schint_cell(ex$cell, "GENE_A", "donor", geno = ex$geno,
                    n_grm = 3, grm_by = "maf", context = "state")
fit3$summary

# or supply GRMs yourself (e.g. from GCTA via read_grm_bin())
K <- make_grm(ex$geno)
fitK <- schint_cell(ex$cell, "GENE_A", "donor", grm = K, context = "state")
```

`geno` can also be a list of matrices (one GRM each, e.g. SNPs by annotation).

## Sample-level data: G x cell type

```{r sample}
fs <- schint_sample(ex$sample, y = "GENE_A", id = "donor", celltype = "celltype",
                    geno = ex$geno, covariates = c("age", "sex", "pc1", "pc2"),
                    jackknife = TRUE)
fs
```

`gxc = FALSE` fits shared genetics only; `cat_mode = "per_level"` estimates a separate
genetic variance for every cell type.

## Perturbation heritability

`schint_pert()` replaces the genetic kernel by "same perturbation". Several genes are
fitted at once and the interaction can be a cell state, a cell type, or both.

```{r pert}
data(schint_pert_example)
fp <- schint_pert(schint_pert_example, y = c("GENE_A", "GENE_B", "GENE_C"),
                  perturb = "perturb", context = c("state", "celltype"),
                  covariates = c("pc1", "pc2"), control = "NT")
fp$summary[fp$summary$term %in% c("P", "PxC:state", "PxC:celltype"), ]
```

Use `perturb_group = "perturb_group"` to estimate separate components for groups of
perturbations.

## Using your own data

1. Genotypes of the cis window of one gene (`read_plink()` + `select_cis_snps()`), as a
   donors x SNPs matrix whose row names match the donor column of your expression table.
2. One row per cell (or donor x cell type) with expression of that gene, the donor ID, the
   context variable(s) and covariates.
3. Call `schint_cell()` / `schint_sample()` once per gene.

Standard errors use a delete-donor (or `n_blocks` blocks of donors) jackknife.
