---
title: "Input formats and outputs"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(scHINT)
data(schint_example)
ex <- schint_example
```

## Inputs

scHINT fits **one gene at a time**: you supply that gene's expression, cell/sample
metadata and cis-genetic information.

### Expression and metadata (`data`)

A data frame (or tibble) with one row per observation.

| model | one row is | required columns |
|---|---|---|
| `schint_cell()` | a cell (of one cell type) | expression `y`, donor `id`, optional `context`, `covariates` |
| `schint_sample()` | a donor × cell-type pseudobulk | expression `y`, donor `id`, `celltype`, optional `covariates` |
| `schint_pert()` | a cell | expression `y` (one or several genes), `perturb`, optional `context` (one cell-state or cell-type column), `covariates` |

```{r cell-head}
head(ex$cell, 3)
```

* **Expression** should be a normalized (e.g. log-normalized, depth-corrected) numeric
  value; it is centred and scaled internally. Non-finite values drop the row.
* **Context** columns: numeric = continuous (e.g. the leading principal components
  computed *within* a cell type, pseudotime, an activation score); factor/character =
  categorical (cell type, subtype, treatment). Numeric context is standardized.
* **Covariates** enter as main effects only. Use cell-level covariates (library size,
  PCs of expression, batch) and donor-level covariates (age, sex, genetic PCs).
  Categorical covariates are one-hot encoded as a "same level" component.
* Donor IDs must match the genetic input exactly (character comparison); donors without
  genotypes and rows with missing values are dropped with a message.

### Genetic input (`geno` **or** `grm`)

| argument | meaning |
|---|---|
| `geno` | donors × SNPs dosage matrix (0/1/2, `NA` allowed), **row names = donor IDs** -- the cis SNPs of the gene; one GRM is built from all SNPs |
| `grm` | a pre-computed donor × donor GRM with donor IDs as dimnames, or a (named) list of GRMs for a multi-GRM model |

With `geno` one GRM is built from all supplied cis SNPs; for several GRMs build them yourself (for
example `lapply(split_snps(geno, 3), make_grm)`) and pass the list to `grm`. Select SNPs within ±500 kb of the gene
(`select_cis_snps()`); genotypes are standardized by allele frequency and the GRM is
$XX^\top/p$. GRMs computed by GCTA can be loaded with `read_grm_bin()`. PLINK files are
read with `read_plink()` (needs the `BEDMatrix` package).

### Perturbation input

`perturb` is a column with the perturbation (target) of each cell. Remove non-targeting
controls with `control = "NT"` (they would otherwise form a large "perturbation");
`min_cells` drops perturbations with too few cells.

## Outputs

### `schint_cell()` and `schint_sample()`

Both return an object of class `"schint"`:

```{r out-cell}
fit <- schint_cell(ex$cell, "GENE_A", "donor", geno = ex$geno, context = "state",
                   covariates = c("pc1", "pc2", "age", "sex", "batch"),
                   jackknife = TRUE, n_blocks = 50)
names(fit)
```

**`fit$summary`** -- grouped variance components (what you usually report):

```{r out-summary}
fit$summary
```

| column | meaning |
|---|---|
| `term` | `G`, `GxC` (sum over context variables), `G_total` = G + GxC, `I`, `context`, `covariate`, `total_explained`; with several GRMs also one row per GRM (`G1`, `G1:C`, ...); with several context variables also `GxC:<variable>` |
| `variance` | variance contribution on the standardized-expression scale |
| `h2` | population-level heritability (cell-level residual removed from the denominator; see [Model](model.html)) |
| `se`, `se_h2` | jackknife standard errors (only with `jackknife = TRUE`) |

**`fit$coefficients`** -- one row per HE-regression term:

```{r out-coef}
fit$coefficients
```

`estimate` is the regression coefficient (variance component of the standardized
variable), `variance` = `estimate` × the average diagonal of its kernel (what the term
contributes to the variance of expression, which is 1). With jackknife: `se`, `z`, `p` (normal approximation) and `se_variance`. `Intercept` is the pair-mean and carries no variance.

Other elements: `n_obs`, `n_donors`, `XtX`/`Xty` (sufficient statistics),
`jackknife$estimates` (leave-block-out estimates, blocks × components), `call`.

### `schint_pert()`

A `"schint_pert"` object with long-format `coefficients` and `summary` tables with an
extra `gene` column (terms: `P`, `PxC`, `P_total`, `context`, `covariate`,
`total_explained`).

### Extracting results

```{r extract}
coef(fit)                      # named vector of regression coefficients
subset(fit$summary, term %in% c("G", "GxC"), select = c(term, h2, se_h2))
```
