---
title: "Simulation and validation"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6.5, fig.height = 3.6)
library(scHINT)
set.seed(1)
```

This page shows how to simulate data with known variance components and how to check that
the estimators recover them. The simulators used for the package example data are
exported: `simulate_geno()`, `simulate_schint()` and `simulate_pert()`.

## Simulated architecture

For each gene, the cell-level generator draws

$$
y_{im}=\sqrt{h_G^2}\,\tilde g_i+\sqrt{h^2_{G\times C}}\,\tilde g'_i\,c_{im}
+\sqrt{h_I^2}\,e_i+\sqrt{h_o^2}\,\tilde o_{im}+\sqrt{h_\varepsilon^2}\,\varepsilon_{im},
$$

with $\tilde g=Xb$, $\tilde g'=Xb'$ polygenic scores from causal SNPs (standardized to unit
variance across donors), $c_{im}$ a standard-normal cell state and the remaining effects
independent. Variances sum to one.

## Bias and variance across replicates

```{r sim, cache = FALSE}
geno <- simulate_geno(400, 250, seed = 11)
spec <- list(GENE = list(g = 0.20, gxc = 0.15, i = 0.10, cov = 0.05))
reps <- 30
est <- t(sapply(seq_len(reps), function(r) {
  s <- simulate_schint(geno, genes = spec, cells = c(20, 40), seed = 100 + r)
  f <- schint_cell(s$cell, "GENE", "donor", geno = geno, context = "state",
                   covariates = c("pc1", "pc2", "age", "sex", "batch", "subtype"))
  setNames(f$summary$variance[match(c("G", "GxC", "I"), f$summary$term)], c("G", "GxC", "I"))
}))
round(rbind(truth = c(G = .20, GxC = .15, I = .10), mean = colMeans(est), sd = apply(est, 2, sd)), 3)
```

```{r sim-plot, echo = FALSE}
op <- par(mfrow = c(1, 3), mar = c(3, 4, 2, 1))
truth <- c(G = .20, GxC = .15, I = .10)
for (k in colnames(est)) {
  boxplot(est[, k], main = k, ylab = "estimate"); abline(h = truth[k], col = "red", lwd = 2)
}
par(op)
```

The red line is the generating value. (The *realized* genetic variance of a random
polygenic score varies around its nominal value; in the cis window the GRM captures only
part of it, so estimates scatter around, rather than exactly at, the nominal heritability.)

## Calibration of the jackknife

```{r calib, cache = FALSE}
z <- t(sapply(seq_len(20), function(r) {
  s <- simulate_schint(geno, genes = list(GENE = list(g = 0.2, gxc = 0, i = 0.1, cov = 0.05)),
                       cells = c(20, 40), seed = 500 + r)
  f <- schint_cell(s$cell, "GENE", "donor", geno = geno, context = "state", jackknife = TRUE)
  f$coefficients[f$coefficients$component == "G:state", c("estimate", "se")]
}))
z <- data.frame(estimate = unlist(z[, 1]), se = unlist(z[, 2]))
c(mean_estimate = mean(z$estimate), sd_of_estimates = sd(z$estimate), mean_jackknife_se = mean(z$se))
```

With no G×Cell-State effect the average estimate is close to zero, and the average jackknife SE
is of the same size as the empirical standard deviation of the estimates.

## Simulating your own scenarios

```r
sim <- simulate_schint(
  geno,
  genes = list(MYGENE = list(g = 0.1, gxc = 0.1, i = 0.05, cov = 0.05)),  # rest is noise
  n_causal = 20, cells = c(30, 60), n_celltypes = 6, seed = 1
)
sim$cell      # cell-level data (one cell type)
sim$sample    # donor x cell-type pseudobulk
sim$truth     # generating variances
```

Use your own genotype matrix to retain real allele-frequency and LD structure.
