---
title: "Tutorial: G×Cell-Type heritability (pseudobulk)"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(scHINT)
data(schint_example)
ex <- schint_example
```

When expression is available as pseudobulk per donor and cell type, `schint_sample()` asks
how much genetic variance is **shared across cell types** (`G`) and how much is
**cell-type specific** (`GxC`).

```{r data}
head(ex$sample, 3)
table(ex$sample$celltype)
```

## Fit

```{r fit}
fs <- schint_sample(ex$sample, y = "GENE_A", id = "donor", celltype = "celltype",
                    geno = ex$geno, covariates = c("age", "sex", "pc1", "pc2"),
                    jackknife = TRUE)
fs
```

* `G` - genetic variance shared by all cell types.
* `GxC` - genetic variance specific to a cell type (pairs from the same type).
* `I` - donor effect shared across cell types, not genetic.
* `context` - variance of cell-type mean differences.


## Shared-genetics model only

```{r shared}
schint_sample(ex$sample, "GENE_A", "donor", "celltype", geno = ex$geno, gxc = FALSE)$summary
```

## A separate genetic variance per cell type

```{r per-level}
fl <- schint_sample(ex$sample, "GENE_A", "donor", "celltype", geno = ex$geno,
                    cat_mode = "per_level", jackknife = TRUE)
subset(fl$coefficients, group == "GxC", select = c(component, variance, se_variance))
```

## Genes without cell-type specificity

```{r geneB}
schint_sample(ex$sample, "GENE_B", "donor", "celltype", geno = ex$geno,
              jackknife = TRUE)$summary[1:3, c("term", "variance", "se")]
```

## Notes

* The data should contain one row per donor × cell type. Several rows per pair trigger a
  warning (use `schint_cell()` for cell-level data).
* Donors need not have all cell types; missing pairs are simply absent.
* Pseudobulk values should be normalized and, if desired, residualized for donor-level
  covariates before calling, or the covariates passed in `covariates`.
