---
title: "Tutorial: G×Cell-State heritability"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6, fig.height = 4)
library(scHINT)
data(schint_example)
ex <- schint_example
```

This tutorial fits the cell-level model to the simulated example. `ex$cell` holds 12,000
cells of one cell type from 400 donors; a continuous cell `state`, cell covariates
(`pc1`, `pc2`), donor covariates (`age`, `sex`, `batch`) and three genes with known
architecture (`ex$truth`):

```{r truth}
ex$truth
```

* `GENE_A`: main genetic + G×state + donor effect
* `GENE_B`: main genetic + donor effect, no interaction
* `GENE_C`: no genetic effect

## 1. Fit the model

```{r fit}
covs <- c("pc1", "pc2", "age", "sex", "batch")
fitA <- schint_cell(ex$cell, "GENE_A", "donor", geno = ex$geno,
                    context = "state", covariates = covs, jackknife = TRUE)
fitA
```

`G` is the genetic variance present in every cell, `GxC` the genetic variance that
varies with `state`, and `G_total` their sum. `h2` / `h2_upper` are the normalized
heritabilities (see [Model](model.html), section 7).

## 2. Is there G×state heritability?

Compare the interaction estimate and its jackknife standard error across genes:

```{r three-genes}
res <- do.call(rbind, lapply(c("GENE_A", "GENE_B", "GENE_C"), function(g) {
  f <- schint_cell(ex$cell, g, "donor", geno = ex$geno, context = "state",
                   covariates = covs, jackknife = TRUE)
  s <- f$summary
  cbind(gene = g, s[s$term %in% c("G", "GxC"), c("term", "variance", "se")])
}))
res$z <- res$variance / res$se
res
```

`GxC` is clearly non-zero for `GENE_A` only.

```{r plot, echo = FALSE}
op <- par(mar = c(4, 4, 1, 1))
x <- barplot(matrix(res$variance, nrow = 2, dimnames = list(c("G", "GxC"), unique(res$gene))),
             beside = TRUE, col = c("grey60", "firebrick"), ylab = "variance",
             ylim = range(c(0, res$variance + 2 * res$se, res$variance - 2 * res$se)))
arrows(x, res$variance - 2 * res$se, x, res$variance + 2 * res$se, angle = 90, code = 3, length = .03)
legend("topright", c("G", "GxC"), fill = c("grey60", "firebrick"), bty = "n")
par(op)
```

## 3. Several context axes

`context` accepts several columns, each with its own main and G× term. Here the
categorical `subtype` is added as a second context (one pooled "same subtype" component):

```{r multi-context}
f2 <- schint_cell(ex$cell, "GENE_A", "donor", geno = ex$geno,
                  context = c("state", "subtype"), covariates = covs)
f2$summary
```

The per-variable terms `GxC:state` and `GxC:subtype` split the interaction. With
`cat_mode = "per_level"` a categorical context gets a separate genetic variance per
level.

## 4. Main-effect-only covariates versus context

A variable in `context` gets a main effect *and* a G× interaction. A variable in
`covariates` gets a main effect only. To mirror a typical analysis, put the leading
state PCs (say 5) in `context` and further PCs and technical variables in `covariates`.

## 5. Multiple GRMs

```{r grm}
f3 <- schint_cell(ex$cell, "GENE_A", "donor", geno = ex$geno,
                  n_grm = 3, grm_by = "maf", context = "state", covariates = covs)
f3$summary[, c("term", "variance", "h2")]
```

Here the cis SNPs are split into three MAF bins (rarest first), each with its own main
and G×state variance; `G` and `GxC` are the sums over GRMs. Pass a list of genotype
matrices, or of GRMs, to split by any annotation.

## 6. Pre-computed GRMs

```{r precomputed}
K <- make_grm(ex$geno)
all.equal(coef(schint_cell(ex$cell, "GENE_A", "donor", grm = K, context = "state",
                           covariates = covs)),
          coef(fitA))
```

## 7. Jackknife options

`n_blocks` groups donors into blocks (faster for very large cohorts); `seed` controls the
assignment. The per-block estimates are in `fit$jackknife$estimates`, from which the
standard error of any function of the components can be computed.

```{r jk-custom}
jk <- schint_cell(ex$cell, "GENE_A", "donor", geno = ex$geno, context = "state",
                  covariates = covs, jackknife = TRUE, n_blocks = 20)
B  <- jk$jackknife$estimates
ratio <- B[, "G:state"] / B[, "G"]                        # a custom statistic
sqrt((length(ratio) - 1) / length(ratio) * sum((ratio - mean(ratio))^2))
```
