---
title: "Tutorial: perturbation heritability"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(scHINT)
data(schint_pert_example)
d <- schint_pert_example
```

`schint_pert()` asks, for each gene, how much expression variance is attributable to the
**perturbation** a cell received, and how much of that effect depends on the **cell state**
and/or **cell type**. The genetic kernel is replaced by "same perturbation".

```{r data}
head(d, 3)
table(d$perturb_group)
```

`NT` marks non-targeting control cells. `GENE_A` has perturbation × state variance,
`GENE_B` perturbation × cell-type variance, `GENE_C` a perturbation effect only and
`GENE_D` none; group3 perturbations have no effect.

## One interaction at a time

The perturbation can interact with **either** a continuous cell state **or** a categorical
cell type in one model (`context` takes a single column); to study both, fit the model
twice. Several genes can be fitted at once, and the expensive part (`X'X`) is computed
only once; each additional gene costs a few quadratic forms.

```{r fit}
genes <- c("GENE_A", "GENE_B", "GENE_C", "GENE_D")
fit_state <- schint_pert(d, y = genes, perturb = "perturb", context = "state",
                         covariates = c("pc1", "pc2"), control = "NT", jackknife = TRUE)
fit_type  <- schint_pert(d, y = genes, perturb = "perturb", context = "celltype",
                         covariates = c("pc1", "pc2"), control = "NT", jackknife = TRUE)
pick <- function(f, label) {
  s <- subset(f$summary, term %in% c("P", "PxC"), select = c(gene, term, variance, se))
  s$interaction <- label
  s$z <- round(s$variance / s$se, 1)
  s
}
rbind(pick(fit_state, "P x cell state"), pick(fit_type, "P x cell type"))
```

`P` is the perturbation variance shared across contexts, and `PxC` the variance of the
perturbation response along the chosen axis. `GENE_A` has a clear **P × cell state**
component (z about 3.5), `GENE_B` the largest **P × cell type** component (z about 2),
and `GENE_C`/`GENE_D` neither. In this simulation the cell state partly tracks cell type, so
`GENE_A` also shows a weak cell-type term; with real data, fit both contexts and compare.

## Perturbation groups

If perturbations fall into groups (pathways, complexes, targets of one regulator) each
group can get its own components:

```{r groups}
fg <- schint_pert(d, c("GENE_A", "GENE_B"), "perturb", context = "state",
                  perturb_group = "perturb_group", control = "NT")
subset(fg$summary, grepl("^P_group", term) & gene == "GENE_A",
       select = c(term, variance))
```

`group3` has no simulated effect, `group1` the largest. A named list works too:
`perturb_group = list(complexA = c("PERT01", "PERT02"), complexB = c(...))`.

## Choosing the context

Use `context = "state"` for a continuous axis (e.g. a principal component of the control
cells) and `context = "celltype"` for a categorical one. `context = NULL` fits
perturbation variance only.

## Practical points

* Remove controls (`control`) or give them their own label and exclude them; otherwise
  controls form one very large "perturbation".
* Perturbations with fewer than `min_cells` (default 2) cells are dropped.
* Standard errors use a leave-perturbation-out jackknife (`n_blocks` for fewer blocks).
* Expression should be normalized; each gene is centred and scaled.
