---
title: "Tutorial: perturbation heritability"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(scHINT)
data(schint_pert_example)
d <- schint_pert_example
```

`schint_pert()` asks, for each gene, how much expression variance is attributable to the
**perturbation** a cell received (`P`), and how much of that effect depends on the cellular
context (`PxC`). The genetic kernel is replaced by "same perturbation". The context is
**one variable at a time**: a continuous **cell state** or a categorical **cell type**.

```{r data}
head(d[, c("perturb", "celltype", "state", "pc1", "pc2", "GENE_A")], 3)
table(d$celltype)
```

`NT` marks non-targeting control cells and is removed with `control = "NT"`. `GENE_A` has
perturbation × cell-state variance, `GENE_B` perturbation × cell-type variance, `GENE_C` a
perturbation effect only and `GENE_D` none. Several genes can be fitted in one call; the
expensive part (`X'X`) is computed only once.

```{r genes}
genes <- c("GENE_A", "GENE_B", "GENE_C", "GENE_D")
```

## Example 1: perturbation × cell type (P×CT)

Set `context` to the cell-type column. `PxC` is the variance of the perturbation response
that is specific to a cell type (cells with the same perturbation **and** the same cell type).

```{r pxct}
fit_ct <- schint_pert(d, y = genes, perturb = "perturb", context = "celltype",
                      covariates = c("pc1", "pc2"), control = "NT", jackknife = TRUE)
res_ct <- subset(fit_ct$summary, term %in% c("P", "PxC"), select = c(gene, term, variance, se))
res_ct$z <- round(res_ct$variance / res_ct$se, 1)
res_ct
```

`GENE_B` has the largest P×CT component.

## Example 2: perturbation × cell state (P×CS)

Set `context` to a continuous cell-state column (for example a principal component of the
control cells). `PxC` is the variance of the perturbation response that changes along that
axis.

```{r pxcs}
fit_cs <- schint_pert(d, y = genes, perturb = "perturb", context = "state",
                      covariates = c("pc1", "pc2"), control = "NT", jackknife = TRUE)
res_cs <- subset(fit_cs$summary, term %in% c("P", "PxC"), select = c(gene, term, variance, se))
res_cs$z <- round(res_cs$variance / res_cs$se, 1)
res_cs
```

`GENE_A` has a clear P×CS component, while `GENE_C` (perturbation effect only) and `GENE_D`
(none) do not. In this simulation the cell state partly tracks cell type, so `GENE_A` also
shows a weak cell-type term in Example 1; with real data fit both and compare.

## Practical points

* Remove controls (`control`) or give them their own label and exclude them; otherwise
  controls form one very large "perturbation".
* Perturbations with fewer than `min_cells` (default 2) cells are dropped.
* Standard errors use a leave-perturbation-out jackknife (`n_blocks` for fewer blocks).
* Expression should be normalized; each gene is standardized.
