---
title: "Tutorial: perturbation heritability"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(scHINT)
data(schint_pert_example)
d <- schint_pert_example
```

`schint_pert()` asks, for each gene, how much expression variance is attributable to the
**perturbation** a cell received, and how much of that effect depends on the **cell state**
and/or **cell type**. The genetic kernel is replaced by "same perturbation".

```{r data}
head(d, 3)
table(d$perturb_group)
```

`NT` marks non-targeting control cells. `GENE_A` has perturbation × state variance,
`GENE_B` perturbation × cell-type variance, `GENE_C` a perturbation effect only and
`GENE_D` none; group3 perturbations have no effect.

## Fit several genes at once

```{r fit}
fit <- schint_pert(d, y = c("GENE_A", "GENE_B", "GENE_C", "GENE_D"),
                   perturb = "perturb", context = c("state", "celltype"),
                   covariates = c("pc1", "pc2"), control = "NT", jackknife = TRUE)
keep <- c("P", "PxC:state", "PxC:celltype")
res <- subset(fit$summary, term %in% keep, select = c(gene, term, variance, se))
res$z <- round(res$variance / res$se, 1)
res
```

The expensive part (`X'X`) is computed once; each additional gene costs only a few
quadratic forms. `P` is the perturbation variance shared across cell states, `PxC:state`
the variance of the perturbation response along the state axis and `PxC:celltype` the
cell-type-specific response.

## Perturbation groups

If perturbations fall into groups (pathways, complexes, targets of one regulator) each
group can get its own components:

```{r groups}
fg <- schint_pert(d, c("GENE_A", "GENE_B"), "perturb", context = c("state", "celltype"),
                  perturb_group = "perturb_group", control = "NT")
subset(fg$summary, grepl("^P_group", term) & gene == "GENE_A",
       select = c(term, variance))
```

`group3` has no simulated effect, `group1` the largest. A named list works too:
`perturb_group = list(complexA = c("PERT01", "PERT02"), complexB = c(...))`.

## Choosing the context

Use `context = "state"` for a continuous axis (e.g. principal components of the control
cells), `context = "celltype"` for categorical, or both, as above. `context = NULL` fits
perturbation variance only.

## Practical points

* Remove controls (`control`) or give them their own label and exclude them; otherwise
  controls form one very large "perturbation".
* Perturbations with fewer than `min_cells` (default 2) cells are dropped.
* Standard errors use a leave-perturbation-out jackknife (`n_blocks` for fewer blocks).
* Expression should be normalized; each gene is centred and scaled.
